---
title: "Imputation scheme: linear model vs logistic model"
date: "`r paste0('Last update: ', format(Sys.time(), '%b %d, %Y'))`"    
---

$$
\newcommand{\E}{\text{E}}
$$

```{r setup}
rm(list = ls())
library(ggplot2)
library(dplyr)
theme_set(theme_bw(base_size=15))
set.seed(2020)

# load some gists
source('https://gist.githubusercontent.com/liangyy/43912b3ecab5d10c89f9d4b2669871c9/raw/3ca651cfa53ffccb8422f432561138a46e93710f/my_ggplot_theme.R')

# load my library
source('../code/rlib_simulation.R')
```

# About

Previous attempt focuses on modeling continuous trait where linear model was used ($y = X\beta + \espilon$). 
Now that the actual data is binary traits, I sketch the logistic version of EM at [here](second_idea_outline_logistic.html).
Here, I do simulation to see how logistic model performs as compared to linear model when the outcomes are binary.

# Simulation

To simulation binary trait, I take liability model.
Specifically, $\phi = X\beta+ \epsilon$ and $y = \mathbb{I}_{\phi > \tau}$ with $\tau = \Phi^{-1}(1 - K)$ and $K$ being disease prevalence.

By heritability, I mean the proportion of liability variation that can be explained by $X\beta$.
And the parameters for simulation are as follow.

```{r param}
sample_size = 10000
n_pheno = 30
prior_causal = 0.4
causal_sigma = 1
n_snp = 20
heritabiltiy = 0.05
maf_low = 0.05
maf_high = 0.45
prevalence = 0.10
```

* Sample size $`r sample_size`$.
* Number of phenotypes $`r n_pheno`$.
* Effect size $\beta_{p, k} \sim \pi \delta_0 + (1 - \pi) N(0, \sigma)$ where $\pi = `r 1 - prior_causal`$ and $\sigma^2 = `r causal_sigma`$.
* Allele frequency $maf \sim Unif(`r maf_low`, `r maf_high`)$.
* Number of SNPs $`r n_snp`$.
* Heritabillty $h^2 = `r heritabiltiy`$ for all phenotypes.
* Disease prevalence $K = `r prevalence`$.

```{r simulation}
maf = get_maf(n_snp, maf_low, maf_high)
h_father = sim_hap(sample_size, n_snp, maf)
h_mother = sim_hap(sample_size, n_snp, maf)
h_child = transmit_haplo(h_father, h_mother)
effect_size = matrix(spike_and_slab(n_snp * n_pheno, 1 - prior_causal, causal_sigma), nrow = n_snp, ncol = n_pheno)
y_father = simulate_pheno_liability(h_father, effect_size, heritabiltiy, maf, prevalence)
y_mother = simulate_pheno_liability(h_mother, effect_size, heritabiltiy, maf, prevalence)
```

# Run EM

## Linear model

```{r lin_em}
source('../code/rlib_em_otf.R')
h1 = h_child[[1]]
h2 = h_child[[2]]
df_z = list()
for(np in c(2, 5, 10, 15, 20, 25, 30)) {
  message('Working on num_pheno = ', np)
  o = em_algorithm_otf(y_father[, 1 : np], y_mother[, 1 : np], h1, h2, maxiter = 50)
  df_z[[length(df_z) + 1]] = data.frame(prob_z = o$gamma, num_pheno = np)
}
df_z = do.call(rbind, df_z)
```

## Logistic model

```{r log_em}
source('../code/em_logistic.R')
h1 = h_child[[1]]
h2 = h_child[[2]]
h1_w_inter = cbind(rep(1, nrow(h1)), h1)
h2_w_inter = cbind(rep(1, nrow(h2)), h2)
df_z_logistic = list()
for(np in c(2, 5, 10, 15, 20, 25, 30)) {
  message('Working on num_pheno = ', np)
  o = em_logistic(y_father[, 1 : np], y_mother[, 1 : np], h1_w_inter, h2_w_inter, prior_prob_z = 0.5, maxiter = 50)
  df_z_logistic[[length(df_z_logistic) + 1]] = data.frame(prob_z = o$prob_z, num_pheno = np)
}
df_z_logistic = do.call(rbind, df_z_logistic)
```

# Results

```{r result}
df = rbind(
  df_z %>% mutate(type = 'linear'),
  df_z_logistic %>% mutate(type = 'logistic')
)
```

```{r vis}
df %>% ggplot() + geom_boxplot(aes(x = factor(num_pheno), y = prob_z, color = type))
tmp_ = df %>% group_by(num_pheno, type) %>% summarize(S_bar = mean(prob_z ^ 2 + (1 - prob_z) ^ 2), gamma_bar = mean(prob_z)) %>% mutate(power_ratio = gamma_bar / sqrt(S_bar))  
tmp_ %>% ggplot() + geom_point(aes(x = factor(num_pheno), y = power_ratio, color = type), position = position_dodge(1))
tmp_ %>% ggplot() + geom_point(aes(x = factor(num_pheno), y = gamma_bar, color = type), position = position_dodge(1))
  
```

