---
title: "For paper: visualizing results on Framingham data"
date: "`r paste0('Last update: ', format(Sys.time(), '%b %d, %Y'))`"    
---

$$
\newcommand{\E}{\text{E}}
\newcommand{\diag}{\text{diag}}
$$

```{r setup}
rm(list = ls())
library(dplyr)
library(ggplot2)
library(data.table)
theme_set(theme_bw(base_size=15))  
bin_prob_z = function(x, cuts = c(0.1, 0.4, 0.6, 0.9)) {
  left_most = cuts[1]
  left_mid = cuts[2]
  right_mid = cuts[3]
  right_most = cuts[4]
  o = rep('', length(x))
  o[x < left_most] = 'strong error'
  o[x >= left_most & x < left_mid] = 'mild error'
  o[x >= left_mid & x <= right_mid] = 'not sure'
  o[x > right_mid & x <= right_most] = 'mild correct'
  o[x > right_most] = 'strong correct'
  o = factor(o, levels = c('strong error', 'mild error', 'not sure', 'mild correct', 'strong correct'))
  o
}
get_ngene = function(logfile) {
  e = fread(cmd = paste0('cat ', logfile, ' | grep SUMMARY'), sep = ' ', data.table = FALSE)
  e = e[length(e) - 3]
  as.numeric(e)
}
power_calc = function(gamma) {
  mean_z = mean(gamma)
  mean_s = mean( gamma ^ 2 + (1 - gamma) ^ 2)
  mean_z / sqrt(mean_s)
}

# load some gists
source('https://gist.githubusercontent.com/liangyy/43912b3ecab5d10c89f9d4b2669871c9/raw/3ca651cfa53ffccb8422f432561138a46e93710f/my_ggplot_theme.R')
```

# About

The effort to analyze Framingham data is motivated by the failure on UK Biobank data where we have a tiny set of parental phenotypes to start with.
Also, simulation results indicate that the more parental phenotypes the greater imputation power would be.
So, we decided to dig into Framingham data where we have some child/father/mother tuple with both transcriptome and genotype available.

The code for this task is at `scripts/framingham_detour/`.

# Haplotype origin inferred from genotypes in family

```{r}
chroms = 1 : 22
df_from_geno = list()
for(chrom in chroms) {
  df = fread(cmd = paste0('zcat < ~/Desktop/tmp/haplotype-po/framingham/infer_ho_from_genotype/from_extracted_pedigree.subset.chr', chrom, '.ho_from_genotype.tsv.gz'), header = TRUE, sep = '\t', data.table = FALSE)
  df_from_geno[[length(df_from_geno) + 1]] = df %>% mutate(chr = chrom)
}
df_from_geno = do.call(rbind, df_from_geno)
```

```{r}
df_from_geno %>% summarize(mean(prob_z))
```

So, the crude approach of assigning haplotype origin indicates that haplotype 1 of child is from father for all pedigrees.

To further verify the result, we look at the observed kinship.
The observed kinship between child's haplotype and parents' genotypes, which is defined as follow with standardized haplotype.

* For diploid (father and mother): $\widetilde{H} = \frac{H^1}{\sqrt{2}} + \frac{H^2}{\sqrt{2}}$ (So that $\widetilde{H}$ has mean 0 and variance 1)
* For haploid (child), $\widetilde{H} = H$.
* Observed kinship is defined as 
    - Between child haplotype and parent genotype: $K_{1,p} = \frac{1}{K} \sum_k H_k^1 \widetilde{H}_k^p$
    - Between parents: $K_{p, q} = \frac{1}{K} \sum_i \widetilde{H}_k^p \widetilde{H}_k^q$
* So that, the expected value of kinship is

$$\begin{aligned}
  \E(K_{1, p}) &= \frac{1}{K} \sum_k \frac{1}{\sqrt{2}}[~ \E(H_k^1 H_k^{p,1}) + \E(H_k^1 H_k^{p,2}) ~] \\
  &= \frac{1}{\sqrt{2}} \\
  \E(K_{p, q}) &= \begin{cases}
    \frac{1}{K} \sum_k \frac{1}{2} [~ \E((H_k^{p, 1})^2) + \E((H_k^2{p, 1})^2) ~] = 1 & \text{, if $p = q$} \\
    0 & \text{, otherwise}
  \end{cases}
\end{aligned}$$

```{r}
df_geno_cor = list()
for(chrom in chroms) {
  df = fread(cmd = paste0('zcat < ~/Desktop/tmp/haplotype-po/framingham/infer_ho_from_genotype/from_extracted_pedigree.subset.chr', chrom, '.child_parents_cor.tsv.gz'), header = TRUE, sep = '\t', data.table = FALSE)
  df_geno_cor[[length(df_geno_cor) + 1]] = df %>% mutate(chr = chrom)
}
df_geno_cor = do.call(rbind, df_geno_cor)
```

```{r, fig.width=10, fig.height=10}
p = df_geno_cor %>% reshape2::melt(id.vars = c('individual_id', 'chr')) %>% 
  ggplot() + 
  geom_boxplot(aes(x = variable, y = value)) + 
  facet_wrap(~chr) + 
  theme(axis.text.x = element_text(angle = 90)) + 
  geom_hline(yintercept = 1 / sqrt(2), linetype = 2, color = 'gray') + 
  th2 + theme(axis.title.x = element_blank()) + 
  ylab('Genetic relatedness'); p
ggsave('../analysis_output/paper_framingham_kinship.png', p, width = 10, height = 10)
```

So, we have high confidence to say that "haplotype 1 of child comes from father for all families".

# Load results

Imputation on full data.


```{r}
models = c('en', 'dapgw')
df_impute = list()
df_ngene = list()
for(chrom in chroms) {
  for(mod in models) {
    df = fread(cmd = paste0('zcat < ~/Desktop/tmp/haplotype-po/framingham/imputation/haplo_impute.', mod, '.chr', chrom, '.tsv.gz'), header = TRUE, sep = '\t', data.table = FALSE)
    logfile = paste0('~/Desktop/tmp/haplotype-po/framingham/imputation/haplo_impute.', mod, '.chr', chrom, '.log')
    df_impute[[length(df_impute) + 1]] = df %>% mutate(chr = chrom, model = mod, fraction = 1)
    df_ngene[[length(df_ngene) + 1]] = data.frame(chr = chrom, model = mod, fraction = 1, ngene = get_ngene(logfile))
  }
}
# df_impute = do.call(rbind, df_impute)
```

Imputation on downsampled data.

```{r}
fractions = 1 : 9 / 10
# df_impute = list()
for(chrom in chroms) {
  for(mod in models) {
    for(frac in fractions) {
      df = fread(cmd = paste0('zcat < ~/Desktop/tmp/haplotype-po/framingham/imputation/haplo_impute.downsample_', frac, '.', mod, '.chr', chrom, '.tsv.gz'), header = TRUE, sep = '\t', data.table = FALSE)
      logfile = paste0('~/Desktop/tmp/haplotype-po/framingham/imputation/haplo_impute.downsample_', frac, '.', mod, '.chr', chrom, '.log')
      df_impute[[length(df_impute) + 1]] = df %>% mutate(chr = chrom, model = mod, fraction = frac) 
      df_ngene[[length(df_ngene) + 1]] = data.frame(chr = chrom, model = mod, fraction = frac, ngene = get_ngene(logfile))
    }
  }
}
df_impute = do.call(rbind, df_impute)
df_impute = df_impute %>% mutate(binned = bin_prob_z(prob_z))
df_ngene = do.call(rbind, df_ngene) %>% mutate(label = paste0('chr', chr, ' ', '\n ngene = ', ngene)) %>% mutate(label = factor(label, levels = unique(label[order(chr)])))
```

## $\gamma$ from imputation

```{r, fig.width=10, fig.height=10}
p = df_impute %>% filter(fraction == 1) %>% left_join(df_ngene %>% filter(fraction == 1), by = c('chr', 'model', 'fraction')) %>% filter(model == 'en', fraction == 1) %>% 
  ggplot() + 
  geom_histogram(aes(prob_z), binwidth = 0.01) + 
  facet_wrap(~label) + 
  xlab(expression(paste('Imputation output ', gamma))) +
  th2 +
  theme(axis.title.y = element_blank()); p
ggsave('../analysis_output/paper_framingham_en.png', p, width = 10, height = 10)
```

```{r, fig.width=10, fig.height=10}
p = df_impute %>% filter(fraction == 1) %>% left_join(df_ngene %>% filter(fraction == 1), by = c('chr', 'model', 'fraction')) %>% filter(model == 'dapgw', fraction == 1) %>% 
  ggplot() + 
  geom_histogram(aes(prob_z), binwidth = 0.01) + 
  facet_wrap(~label) + 
  xlab(expression(paste('Imputation output ', gamma))) +
  th2 +
  theme(axis.title.y = element_blank()); p
ggsave('../analysis_output/paper_framingham_dapgw.png', p, width = 10, height = 10)
```

## Downsample analysis 

Downsample the number of genes by keeping a fraction of the total genes randomly.

```{r, fig.width=10, fig.height=10}
p = df_impute %>% filter(model == 'en') %>% 
  left_join(df_ngene %>% filter(fraction == 1) %>% select(-fraction), by = c('chr', 'model')) %>% 
  ggplot() + geom_bar(aes(binned, fill = fraction, group = fraction), position = 'dodge') + 
  facet_wrap(~label) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
  xlab(expression(paste('Imputation output ', gamma, ' in bins'))) +
  th2 + 
  theme(axis.title.y = element_blank()) +
  theme(legend.title = element_blank()) + 
  scale_fill_gradient(low = "yellow", high = "red") +
  theme(legend.position = c(0.7, 0.03), legend.direction = "horizontal", legend.key.width = unit(2.5,"cm") ); p
ggsave('../analysis_output/paper_framingham_en_downsample.png', p, width = 10, height = 10)
```

```{r, fig.width=10, fig.height=10}
p = df_impute %>% filter(model == 'dapgw') %>% 
  left_join(df_ngene %>% filter(fraction == 1) %>% select(-fraction), by = c('chr', 'model')) %>% 
  ggplot() + geom_bar(aes(binned, fill = fraction, group = fraction), position = 'dodge') + 
  facet_wrap(~label) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
  xlab(expression(paste('Imputation output ', gamma, ' in bins'))) +
  th2 + 
  theme(axis.title.y = element_blank()) +
  theme(legend.title = element_blank()) + 
  scale_fill_gradient(low = "yellow", high = "red") +
  theme(legend.position = c(0.7, 0.03), legend.direction = "horizontal", legend.key.width = unit(2.5,"cm") ); p
ggsave('../analysis_output/paper_framingham_dapgw_downsample.png', p, width = 10, height = 10)
```

## Power of GWAS with imputed haplotype

```{r, fig.width=10, fig.height=10}
label_chr = function(x) {
  paste0('chr', x)
}
p = df_impute %>% group_by(model, chr, fraction) %>% 
  summarize(power = power_calc(prob_z)) %>% ungroup() %>% 
  left_join(df_ngene, by = c('chr', 'model', 'fraction')) %>% 
  mutate(power_gain_relative_to_avg = power * sqrt(2)) %>% 
  ggplot() + 
  geom_hline(yintercept = 1, linetype = 1, color = 'gray') + 
  geom_vline(xintercept = 10, linetype = 2, color = 'gray') + 
  geom_point(aes(x = ngene, y = power_gain_relative_to_avg, color = model)) + 
  facet_wrap(~chr, labeller = as_labeller(label_chr)) + 
  scale_x_log10() +
  th2 +
  xlab('Number of genes in the imputation') +
  ylab('Expected power increase relative to GWAX') +
  theme(legend.position = c(0.7, 0.03), legend.title = element_blank(), legend.direction = "horizontal", legend.text = element_text(size = 15)) +
  scale_color_discrete(breaks = c("en", "dapgw"), labels = c("EN", 'EN DAPG')); p
ggsave('../analysis_output/paper_framingham_power.png', p, width = 10, height = 10)
```

