---
title: "Sanity check on first pass effort"
date: "`r paste0('Last update: ', format(Sys.time(), '%b %d, %Y'))`"    
bibliography: ref.bib
---


```{r setup}
rm(list = ls())
library(ggplot2)
library(dplyr)
library(patchwork)
library(data.table)
library(reticulate)
library(pander)
options(datatable.fread.datatable = FALSE)
np <- import("numpy")
set.seed(2020)
source('../code/rlib_misc.R')
```

# About

I've implemented a pipeline doing the haplotype imputation based GWAS which includes three major scripts:

* **Calculate PRS**: `scripts/prs/naive_prs.py`
* **Imputation**: `scripts/haplotype_imputation/impute_parent_of_origin.py`
* **Run GWAS**: `scripts/logistic_gpu/run_haplo_logistic_solver.py`

As first pass effort, I run GWAS on paternal phenotype with imputed haplotype (Alzheimer's disease, AD).
The result is directly comparable to the paternal GWAS  on AD reported in @marioni:2018.
I tested on chromosome 16 and 19.

Two imputation (labeled as `imp`) approaches are tested: 1) sex-specific EM (labeled as `em`); 2) non sex-specific EM (as `em_deg`). Besides, I include $\Pr(Z) = 0.5$ as `avg`, and `flip` which flip the imputation by using $1 - Pr(Z)$.

Besides, I also run on heart disease (HD) on the side, though I don't have an existing GWAS to compare to.

# Load results

Some functions.

```{r load_func}
to_df = function(mat) {
  df = as.data.frame(mat)
  colnames(df) = c('beta', 'se', 'convergence')
  return(df)
}
read_from_npy = function(filename, pos, trait_idx) {
  check_imputer = np$load(filename)
  df_trait2 = rbind(
    to_df(check_imputer[1, trait_idx, , ]) %>% 
      mutate(snpid = 1 : nrow(.), position = pos, impute = 'imp'),
    to_df(check_imputer[2, trait_idx, , ]) %>% 
      mutate(snpid = 1 : nrow(.), position = pos, impute = 'avg'),
    to_df(check_imputer[3, trait_idx, , ]) %>% 
      mutate(snpid = 1 : nrow(.), position = pos, impute = 'flip')
  )
  # extract snps that converge
  df_trait2 = df_trait2 %>% filter(convergence == 1, !is.na(se), se > 0) %>% filter(abs(beta) < 100, se < 5)
  df_trait2
}
signed_beta = function(beta, allele_ids, a1, a2) {
  ref = unlist(lapply(strsplit(allele_ids, ','), function(x) {x[1]}))
  alt = unlist(lapply(strsplit(allele_ids, ','), function(x) {x[2]}))
  obeta = rep(NA, length(beta))
  obeta[ref == a2 & alt == a1] = beta[ref == a2 & alt == a1]
  obeta[ref == a1 & alt == a2] = -beta[ref == a1 & alt == a2]
  return(obeta)
}
get_id = function(id) {
  unlist(lapply(strsplit(id, ':'), function(x) {
    paste0(x[1], ':', x[2])
  }))
}
```

```{r load}
snp_map = fread('zcat < ~/Desktop/tmp/haplotype-po/snp_map_for_neale_lab_gwas.with_sign.tsv.gz', header = TRUE, sep = '\t')
gwas_father = fread('~/Downloads/2_UKB_AD_paternal_summary_output_June2019.txt', sep = ' ')

methods = data.frame(method = c('em', 'em_deg', 'em_otf', 'em_otf_w_covar', 'em_otf_w_covar_hypertension', 'em_per_snp_w_covar_hypertension'), prefix = c('test_sanity_imputer_chr', 'test_sanity_imputer_deg_chr', 'test_sanity_imputer_otf_chr', 'test_sanity_imputer_otf_w_covar_chr', 'test_sanity_imputer_otf_w_covar_hypertension_chr', 'test_sanity_imputer_otf_per_snp_w_covar_hypertension_chr'))
dl = list()
dl[[1]] = list()
dl[[2]] = list()
dl[[3]] = list()
for(chr_num in c(16, 19)) {
  pos = snp_map %>% filter(chrom == chr_num) %>% pull(pos)
  for(m in 1 : nrow(methods)) {
    for(i in 1 : 2) {
      if(!file.exists(paste0('/Users/yanyul/Desktop/tmp/haplotype-po/', methods$prefix[m], chr_num, '.npy'))) {
        next
      }
      df = read_from_npy(
        paste0('/Users/yanyul/Desktop/tmp/haplotype-po/', methods$prefix[m], chr_num, '.npy'),
        pos, 
        i
      )
      df = df %>% mutate(chr = chr_num, method = methods$method[m])
      dl[[i]][[length(dl[[i]]) + 1]] = df
    }
    if(methods$method[m] %in% c('em_otf_w_covar_hypertension', 'em_per_snp_w_covar_hypertension')) {
      if(!file.exists(paste0('/Users/yanyul/Desktop/tmp/haplotype-po/', methods$prefix[m], chr_num, '.npy'))) {
        next
      }
      df = read_from_npy(
        paste0('/Users/yanyul/Desktop/tmp/haplotype-po/', methods$prefix[m], chr_num, '.npy'),
        pos, 
        3
      )
      df = df %>% mutate(chr = chr_num, method = methods$method[m])
      dl[[3]][[length(dl[[3]]) + 1]] = df
    }
  }
}
df = list()
info = c('HD', 'AD', 'HT')
for(i in 1 : length(info)) {
  df[[i]] = do.call(rbind, dl[[i]]) %>% mutate(trait = info[i])
}
df = do.call(rbind, df)
```


# AD

## Prepare data frame

```{r ad}
df_ad = df %>% filter(trait == 'AD')
extract_snp = snp_map %>% select(chrom, pos, allele_ids) %>% filter(chrom %in% c(16, 19)) %>% inner_join(
  gwas_father, 
  by = c('pos' = 'BP', 'chrom' = 'CHR')
)
extract_snp = extract_snp %>% mutate(beta = signed_beta(Beta, allele_ids, A1, A2)) %>% mutate(impute = 'avg', method = 'paper')
df_ad = rbind(
  df_ad %>% select(beta, se, position, impute, chr, method), 
  extract_snp %>% select(beta, SE, pos, impute, chrom, method) %>% rename(se = SE, position = pos, chr = chrom)
)
df_ad %>% group_by(method, impute) %>% summarize(nsnp = n()) %>% pander
```

## Effect size

```{r ad_beta, fig.height=10, fig.width=5}
df_paper = df_ad %>% filter(method == 'paper')
df_paper_plot = df_paper %>% inner_join(df_ad %>% filter(method != 'paper'), by = c('position', 'chr'), suffix = c('_paper', '_other'))
df_paper_plot %>% ggplot() + geom_point(aes(x = beta_paper, y = beta_other), alpha = .2) + facet_grid(method_other~impute_other) + ggtitle('paper beta vs my beta') + geom_abline(slope = 1, intercept = 0, color = 'gray')
```

```{r ad_beta2, fig.height=3, fig.width=5}
# compare em vs em_deg
df_ad_plot = df_ad %>% filter(method %in% c('em', 'em_deg')) %>% reshape2::dcast(position + impute + chr ~ method, value.var = 'beta')
df_ad_plot %>% ggplot() + geom_point(aes(x = em, y = em_deg), alpha = 0.2) + facet_wrap(~impute) + ggtitle('compare em vs em_deg: beta')
df_ad_plot = df_ad %>% filter(method %in% c('em', 'em_deg')) %>% reshape2::dcast(position + impute + chr ~ method, value.var = 'se')
df_ad_plot %>% ggplot() + geom_point(aes(x = em, y = em_deg), alpha = 0.2) + facet_wrap(~impute) + ggtitle('compare em vs em_deg: se')
```

```{r ad_beta2, fig.height=3, fig.width=5}
# compare em vs em_deg
df_ad_plot = df_ad %>% filter(method %in% c('em_otf_w_covar_hypertension', 'em_per_snp_w_covar_hypertension')) %>% reshape2::dcast(position + impute + chr ~ method, value.var = 'beta')
df_ad_plot %>% ggplot() + geom_point(aes(x = em_otf_w_covar_hypertension, y = em_per_snp_w_covar_hypertension), alpha = 0.2) + facet_wrap(~impute) + ggtitle('compare em vs em_deg: beta')
df_ad_plot = df_ad %>% filter(method %in% c('em_otf_w_covar_hypertension', 'em_per_snp_w_covar_hypertension')) %>% reshape2::dcast(position + impute + chr ~ method, value.var = 'se')
df_ad_plot %>% ggplot() + geom_point(aes(x = em_otf_w_covar_hypertension, y = em_per_snp_w_covar_hypertension), alpha = 0.2) + facet_wrap(~impute) + ggtitle('compare em vs em_deg: se')
```

## P-value

```{r ad_qq}
df_ad_qq = df_ad %>% mutate(pval = beta2pval(beta, se)) %>% group_by(method, impute, chr) %>% mutate(p_exp = rank(pval) / (n() + 1))
df_ad_qq %>% ggplot() + geom_point(aes(x = -log(p_exp), y = -log(pval), color = impute)) + facet_grid(chr~method, scales = 'free_y') + ggtitle('QQ-plot by chromsome')
```

```{r ad_pval, fig.height=8, fig.width=6}
df_avg = df_ad_qq %>% filter(impute == 'avg')
df_avg_plot = df_avg %>% inner_join(df_ad_qq %>% filter(impute != 'avg'), by = c('position', 'chr', 'method'), suffix = c('_avg', '_other'))

p1 = df_avg_plot %>% filter(chr == 16) %>% ggplot() + geom_point(aes(x = -log(pval_avg), y = -log(pval_other), color = impute_other), alpha = 0.3) + facet_grid(method~impute_other, scales = 'free') + ggtitle('Avg log p vs my log p') + geom_abline(slope = 1, intercept = 0, color = 'gray')
p2 = df_avg_plot %>% filter(chr == 19) %>% ggplot() + geom_point(aes(x = -log(pval_avg), y = -log(pval_other), color = impute_other), alpha = 0.3) + facet_grid(method~impute_other, scales = 'free') + ggtitle('Avg log p vs my log p') + geom_abline(slope = 1, intercept = 0, color = 'gray') # + coord_cartesian(xlim = c(0,25), ylim = c(0,25))
p1 / p2
```



# HD

```{r hd}
df_hd = df %>% filter(trait == 'HD')
```

## Effect size

```{r hd_beta}
df_em_avg = df_hd %>% filter(method == 'em', impute == 'avg')
df_em_avg_plot = df_em_avg %>% inner_join(df_hd, by = c('position', 'chr'), suffix = c('_em_avg', '_other'))
df_em_avg_plot %>% ggplot() + geom_point(aes(x = beta_em_avg, y = beta_other), alpha = .2) + facet_grid(method_other~impute_other) + ggtitle('em avg beta vs my beta') + geom_abline(slope = 1, intercept = 0, color = 'gray')
df_em_avg_plot %>% ggplot() + geom_point(aes(x = se_em_avg, y = se_other), alpha = .2) + facet_grid(method_other~impute_other) + ggtitle('em avg se vs my se') + geom_abline(slope = 1, intercept = 0, color = 'gray')
```

## P-value

```{r hd_qq}
df_hd_qq = df_hd %>% mutate(pval = beta2pval(beta, se)) %>% group_by(method, impute, chr) %>% mutate(p_exp = rank(pval) / (n() + 1))
df_hd_qq %>% ggplot() + geom_point(aes(x = -log(p_exp), y = -log(pval), color = impute)) + facet_grid(chr~method, scales = 'free_y') + ggtitle('QQ-plot by chromsome')
```

```{r hd_pval, fig.height=8, fig.width=6}
df_avg = df_hd_qq %>% filter(impute == 'avg')
df_avg_plot = df_avg %>% inner_join(df_hd_qq %>% filter(impute != 'avg'), by = c('position', 'chr', 'method'), suffix = c('_avg', '_other'))

p1 = df_avg_plot %>% filter(chr == 16) %>% ggplot() + geom_point(aes(x = -log(pval_avg), y = -log(pval_other), color = impute_other), alpha = 0.3) + facet_grid(method~impute_other, scales = 'free') + ggtitle('Avg log p vs my log p') + geom_abline(slope = 1, intercept = 0, color = 'gray')
p2 = df_avg_plot %>% filter(chr == 19) %>% ggplot() + geom_point(aes(x = -log(pval_avg), y = -log(pval_other), color = impute_other), alpha = 0.3) + facet_grid(method~impute_other, scales = 'free') + ggtitle('Avg log p vs my log p') + geom_abline(slope = 1, intercept = 0, color = 'gray')
p1 / p2
```


# HT

```{r ht}
df_ht = df %>% filter(trait == 'HT') %>% mutate(pval = beta2pval(beta, se))
```

## Neale lab GWAS

```{r neale}
gwas_self = fread('zcat < ~/Desktop/tmp/haplotype-po/20002_1065.gwas.imputed_v3.both_sexes.tsv.bgz | grep "^16:\\|^19:"', header = F, sep = '\t', data.table = F)
gwas_parent = fread('zcat < ~/Desktop/tmp/haplotype-po/20107_8.gwas.imputed_v3.both_sexes.tsv.bgz | grep "^16:\\|^19:"', header = F, sep = '\t', data.table = F)
gwas_self = gwas_self %>% mutate(snp = get_id(V1))
gwas_parent = gwas_parent %>% mutate(snp = get_id(V1))
gwas_self = gwas_self %>% filter(snp %in% unique(paste0(df_ht$chr, ':', df_ht$position)))
gwas_parent = gwas_parent %>% filter(snp %in% unique(paste0(df_ht$chr, ':', df_ht$position)))
gwas_ht = inner_join(gwas_self, gwas_parent, by = 'snp', suffix = c('.self', '.parent'))
gwas_ht %>% filter(V12.self < 0.01, V12.parent < 0.01, !V5.self, !V5.parent) %>% ggplot() + geom_point(aes(x = V9.self, y = V9.parent)) 
gwas_ht %>% filter(V12.self < 0.01, V12.parent < 0.01, !V5.self, !V5.parent) %>% ggplot() + geom_point(aes(x = -log(V12.self), y = -log(V12.parent))) 
gwas_ht_bonferroni = gwas_ht %>% filter(V12.self < 0.01 / n() | V12.parent < 0.01 / n(), !V5.self, !V5.parent)
df_ht %>% filter(paste0(chr, ':', position) %in% gwas_ht_bonferroni$snp) %>% reshape2::dcast(position + chr + method ~ impute, value.var = 'pval') %>% ggplot() + geom_point(aes(x = -log(avg), y = -log(flip), color = 'flip')) + geom_point(aes(x = -log(avg), y = -log(imp), color = 'imp')) + geom_abline(slope = 1, intercept = 0) + facet_grid(method~chr)
```

<!-- ## Effect size -->

<!-- ```{r hd_beta} -->
<!-- df_em_avg = df_ht %>% filter(impute == 'avg') -->
<!-- df_em_avg_plot = df_em_avg %>% inner_join(df_ht, by = c('position', 'chr'), suffix = c('_em_avg', '_other')) -->
<!-- df_em_avg_plot %>% ggplot() + geom_point(aes(x = beta_em_avg, y = beta_other), alpha = .2) + facet_grid(chr~impute_other) + ggtitle('em avg beta vs my beta') + geom_abline(slope = 1, intercept = 0, color = 'gray') -->
<!-- df_em_avg_plot %>% ggplot() + geom_point(aes(x = se_em_avg, y = se_other), alpha = .2) + facet_grid(chr~impute_other) + ggtitle('em avg se vs my se') + geom_abline(slope = 1, intercept = 0, color = 'gray') -->
<!-- ``` -->

## P-value

```{r hd_qq}
df_ht_qq = df_ht %>% group_by(method, impute, chr) %>% mutate(p_exp = rank(pval) / (n() + 1))
df_ht_qq %>% ggplot() + geom_point(aes(x = -log(p_exp), y = -log(pval), color = impute)) + facet_grid(method~chr, scales = 'free_y') + ggtitle('QQ-plot by chromsome')
```

<!-- ```{r hd_pval, fig.height=8, fig.width=6} -->
<!-- df_avg = df_hd_qq %>% filter(impute == 'avg') -->
<!-- df_avg_plot = df_avg %>% inner_join(df_hd_qq %>% filter(impute != 'avg'), by = c('position', 'chr', 'method'), suffix = c('_avg', '_other')) -->

<!-- p1 = df_avg_plot %>% filter(chr == 16) %>% ggplot() + geom_point(aes(x = -log(pval_avg), y = -log(pval_other), color = impute_other), alpha = 0.3) + facet_grid(method~impute_other, scales = 'free') + ggtitle('Avg log p vs my log p') + geom_abline(slope = 1, intercept = 0, color = 'gray') -->
<!-- p2 = df_avg_plot %>% filter(chr == 19) %>% ggplot() + geom_point(aes(x = -log(pval_avg), y = -log(pval_other), color = impute_other), alpha = 0.3) + facet_grid(method~impute_other, scales = 'free') + ggtitle('Avg log p vs my log p') + geom_abline(slope = 1, intercept = 0, color = 'gray') -->
<!-- p1 / p2 -->
<!-- ``` -->


# References
